### 2016-09-29 Projects/DelCo_2014_2015/Reports/Propane_Gas_Fire_Experiments/TeX_Files/Propane_Gas_Fire_Experiments.tex

\headerC{
\flushright{
Joseph M. Willi \\
Craig G. Weinschenk$^*$ \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\vspace*{2\baselineskip}
\textit{$^*$\small Currently with JENSEN HUGHES}
}

@@

\chapter*{\centering Abstract}
\pagenumbering{gobble}
Nine full-scale fire experiments were conducted in two residential-scale structures to study how different fire fighting tactics, specifically horizontal, vertical, and positive pressure ventilation, affect the fire environment within a structure. The fire source for each experiment was provided by three propane burners. Five of the tests were conducted in a structure designed to simulate a single-story residential structure, and four of the tests were conducted in a structure designed to simulate a two-story residential structure. The structures were instrumented with sensors to collect local measurements of temperature, gas velocity, heat flux, and gas concentrations at various locations during the experiments. Different doors and vents were opened and closed during the tests to simulate horizontal and vertical ventilation. A PPV fan was also used during some of the experiments to further affect the ventilation conditions. This report contains detailed descriptions of the setup and procedure used for each of the nine experiments.

@@

The development and behavior of compartment fires, such as a fire inside a residential structure, depend greatly on the ventilation conditions within the compartment. There are a variety of tactics used by firefighters that can affect the ventilation of a fire environment. For example, firefighters may open different doors and windows within a structure or cut holes in a structure's roof with the intention of removing hot gases and smoke from the structure's interior and introducing fresh, cool air into the environment. Additionally, firefighters sometimes employ a tactic known as ``positive pressure attack'' (PPA) or ``positive pressure ventilation'' (PPV) on the fire ground in an attempt to improve the tenability of the fire environment. PPA/PPV involves using a fan aimed at an opening of a structure and is intended to force fresh, exterior air through the opening into the structure and exhaust hot gases and smoke out of the structure through a separate vent.

Nine full-scale fire experiments were performed to study how the tactics described above affect ventilation and the fire environment within a structure. The tests were conducted in two structures designed to replicate typical residential dwellings. During each experiment, propane was provided to three diffusion flame burners. Local measurements of temperature, gas velocity, heat flux, and gas concentrations were collected at various locations throughout the structure while the ventilation within the structure was varied by opening and closing certain doors and vents. During a number of the tests, a PPV fan was used in conjunction with the opening and closing of vents to further affect the ventilation conditions within the fire environment. This report contains in-depth descriptions of the setup and procedure used for each of the nine experiments.

% ======================
% = EXPERIMENTAL SETUP =
% ======================
\chapter{Experimental Setup}
\label{chap:Experimental_Setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center in Sharon Hill, PA. Three propane burners were used to provide the fire source for all experiments, and various sensors were used to collect gas temperature, gas velocity, heat flux, and gas concentration measurements throughout the structure.

@@

The first floor ceiling support of each structure was composed of wood truss joist I-beams (TJIs). Each TJI had a depth of 298~mm (11.75~in) and contained laminated veneer lumber flanges with a cross section of 29~mm (1.13~in) x 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board (OSB) web as shown in Fig.~\ref{fig:TJI}. A layer of 18.3~mm (0.72~in) thick tongue and groove OSB was attached to the top of the TJIs.

@@

\subsection{Measurement Uncertainty}
\label{sec:Uncertainty}
There are different components of uncertainty in the reported length, mass, temperature, heat flux, gas concentration, differential pressure, gas velocity, and heat release rate measurements. Uncertainties are grouped into two categories according to the method used to estimate them. Type A uncertainties are those which are evaluated by statistical methods, and Type B are those which are evaluated by other means~\cite{Taylor&Kuyatt:1994}. Type B analysis of systematic uncertainties involves estimating the upper (+a) and lower (-a) limits for the quantity in question such that the probability that the value would be in the interval ($\pm$a) is essentially 100~\%. After estimating uncertainties by either Type A or B analysis, the uncertainties are combined in quadrature to yield the combined standard uncertainty. Then, the combined standard uncertainty is multiplied by a coverage factor of two, which results in the expanded uncertainty with a 95~\% confidence interval (2$\sigma$). For some of these components, such as the zero and calibration elements, uncertainties are derived from referenced instrument specifications. For other components, referenced research results and past experience with the instruments provided input in the uncertainty determination.

\subsubsection{Compartment Dimensions}
Each length measurement was taken carefully. Length measurements such as the room dimensions and instrumentation array locations were made with a hand held laser measurement device, which has an accuracy of $\pm$6.0~mm (0.25~in) over a range of 0.61~m (2.0~ft) to 15.3~m (50.0~ft)~\cite{StanleyTools}. However, conditions affecting the measurement, such as levelness of the device, yields an estimated uncertainty of $\pm$0.5~\% for measurements in the 2.0~m (6.6~ft) to 10.0~m (32.8~ft) range. Steel measuring tapes with a resolution of $\pm$0.5~mm (0.02~in) were used to locate individual sensors within a measurement array and to measure and position the furniture. The steel measuring tapes were manufactured in compliance with NIST Manual 44, which specifies a tolerance of $\pm$1.6~mm (0.06~in) for 9.1~m (30~ft) tapes and $\pm$6.4~mm (0.25~in) for 30.5~m (100~ft) tapes~\cite{Butcher:2012}. Some issues, such as ``soft'' edges on the upholstered furniture, result in an estimated total expanded uncertainty of $\pm$1.0~\% for length measurements.

\subsubsection{Thermocouples}
The standard uncertainty in the temperature of the thermocouple wire itself as determined by the wire manufacturer is $\pm$2.2$~^{\circ}$C at $277~^{\circ}$C and increases to $\pm$9.5$~^{\circ}$C at $871~^{\circ}$C~\cite{Omega:2004}. The variation of the temperature in the environment surrounding the thermocouple is known to be much greater than that of the wire uncertainty. Maximum percent error has been reported to be as high as 20~\% for upper layer temperatures measured by a 1~mm bare-bead type K thermocouple~\cite{Blevins:1999,Pitts:2003}. Small diameter thermocouples were used during these experiments to limit the impact of radiative heating and cooling. The estimated total expanded uncertainty associated with the temperature measurements is $\pm$15~\%.

\subsubsection{Heat Flux Gauges}
Total heat flux measurements were made with water-cooled Schimidt-Bolter gauges. The manufacturer reports a $\pm$3~\% calibration expanded uncertainty for these devices~\cite{Medtherm:2003}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge is typically $\pm$8~\%~\cite{Pitts:2006}.

\subsubsection{Gas Sampling}
The gas measurement instruments and sampling system used in this series of experiments have demonstrated an expanded (k = 2) relative uncertainty of $\pm$1~\% when compared with span gas volume fractions~\cite{Bundy:2007}. Given the non-uniformities and movement of the fire gas environment and the limited set of sampling points in these experiments, an estimated uncertainty of $\pm$12~\% is associated with the gas concentrations~\cite{Lock:1}.

\subsubsection{Pressure Transducers}
Differential pressure reading uncertainty components were derived from pressure transducer instrument specifications and previous experience with pressure transducers. The transducers were factory calibrated, and the zero and span of each were checked in the laboratory prior to the experiments yielding an accuracy of $\pm$1~\%~\cite{Setra:2002}. The total expanded uncertainty is estimated at 10~\%.

\subsubsection{Bi-Directional Probes}
Bi-directional probe and single thermocouple pairs were used to measure gas velocity. The bi-directional probes used similar pressure transducers as those used for the differential pressure measurements discussed above. A single bare-bead Chromel-Alumel (type K) thermocouple with a 1.0~mm (0.04~in) nominal diameter was co-located with each probe. The thermocouple wire was protected with a 3.2~mm (0.13~in) diameter inconel sheath. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm$0.14 to $\pm$0.22 for bi-directional probes similar to the ones described here~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity measurements is estimated to be $\pm$18~\%.

\subsubsection{Heat Release Rate}
A rotary type gas meter was used to measure the volume of propane provided to the gas burners. The manufacturer quantifies the maximum error of the meter as $\pm$2~\%~\cite{Romet:2014}. A volumetric flow rate was calculated from the gas meter volume readings and used in conjunction with the heat of combustion of propane to calculate the heat release rate of the fire. The total expanded uncertainty for the heat release rate obtained from this method is estimated to be $\pm12$~\%.

@@

\chapter{Acknowledgments}
\label{chap:acknowledgments}
The authors would like to thank Daniel Madrzykowski (currently with UL FSRI), Roy McLane (currently with Thermal Fabrication), Kristopher Overholt (currently with Continuum Analytics), and Keith Stakes (currently with UL FSRI) from NIST for their contributions in preparing and conducting these experiments. These experiments could not have been conducted without them.

>>>>

\headerC{
\flushright{
Joseph M. Willi \\
Craig G. Weinschenk \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

}

@@

\chapter*{\centering Abstract}
\pagenumbering{gobble}
Nine full-scale fire experiments were conducted in two residential-scale structures to measure the effect of passive and forced ventilation on the fire conditions. The fires were fueled by three propane burners. Five of the experiments were conducted in a single-story structure, and four were conducted in a two-story structure. The structures were instrumented to measure temperature, gas velocity, heat flux, and oxygen, carbon dioxide and carbon monoxide concentrations. Various doors and vents were opened and closed during the experiments to provide passive horizontal and vertical ventilation. A fan was used to provide forced ventilation. This report contains detailed descriptions of the setup and procedure used for each of the nine experiments.

@@

The development and behavior of compartment fires, such as a fire inside a residential structure, depend greatly on the ventilation conditions within the compartment. There are a variety of tactics used by firefighters that can affect the ventilation of a fire environment. For example, firefighters may open different doors and windows within a structure or cut holes in the roof to remove hot gases and smoke. Additionally, firefighters sometimes employ a tactic known as ``positive pressure attack'' (PPA) or ``positive pressure ventilation'' (PPV) in an attempt to improve the tenability of the fire environment. PPA/PPV involves using a fan aimed at an opening of a structure and is intended to force fresh air into the structure and force hot gases and smoke out through a separate exterior vent.

Nine full-scale fire experiments were performed to study how the tactics described above affect ventilation and the fire environment within a structure. The experiments were conducted in two structures designed to replicate typical residential dwellings. During each experiment, propane was provided to three diffusion flame burners. Local measurements of temperature, gas velocity, heat flux, and gas concentrations were collected at various locations throughout the structure while the ventilation within the structure was varied by opening and closing certain doors and vents. During a number of the tests, a PPV fan was used in conjunction with the opening and closing of vents to further affect the ventilation conditions within the fire environment. This report contains in-depth descriptions of the setup and procedure used for each of the nine experiments.

% ======================
% = EXPERIMENTAL SETUP =
% ======================
\chapter{Experimental Setup}
\label{chap:Experimental_Setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center in Sharon Hill, Pennsylvania. Three propane burners were used to fuel the fire for all experiments, and various sensors were used to collect gas temperature, gas velocity, heat flux, and gas concentration measurements throughout the structure.

@@

The first floor ceiling support of each structure was composed of wood truss joist I-beams (TJIs). Each TJI had a depth of 298~mm (11.75~in) and contained laminated veneer lumber flanges with a cross section of 29~mm (1.13~in) by 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board (OSB) web as shown in Fig.~\ref{fig:TJI}. A layer of 18.3~mm (0.72~in) thick tongue and groove OSB was attached to the top of the TJIs.

@@

\subsection{Measurement Uncertainty}
\label{sec:Uncertainty}
This section lists the uncertainties in the reported length, mass, temperature, heat flux, gas concentration, gas velocity, and heat release rate measurements. Uncertainty estimates are based either on manufacturer literature or analyses performed by others for similar measurement devices and techniques. In accordance with NIST guidelines~\cite{Taylor&Kuyatt:1994}, measurement accuracy is reported as an {\em expanded uncertainty} or 95~\% ($2\,\sigma$) confidence interval. Most manufactures express accuracy in terms of a {\em standard  uncertainty} or 68~\% ($1\,\sigma$) confidence interval.

\subsubsection{Compartment Dimensions}
Room dimensions and instrumentation array locations were made with a hand held laser measurement device with a standard uncertainty of $\pm$6.0~mm (0.25~in) over a range of 0.6~m (2.0~ft) to 15~m (50.0~ft)~\cite{StanleyTools}. Steel measuring tapes with a resolution of $\pm$0.5~mm (0.02~in) were used to locate measurement devices. The steel measuring tapes were manufactured in compliance with NIST Manual 44, which specifies a tolerance of $\pm$1.6~mm (0.06~in) for 9.1~m (30~ft) tapes and $\pm$6.4~mm (0.25~in) for 30.5~m (100~ft) tapes~\cite{Butcher:2012}. These uncertainties are all well within the precision of the reported dimensions, which are typically rounded to the nearest 10~cm (4~in).

\subsubsection{Thermocouples}
The standard uncertainty in the temperature of the thermocouple wire itself as determined by the wire manufacturer is $\pm$2.2$~^{\circ}$C at $277~^{\circ}$C and increases to $\pm$9.5$~^{\circ}$C at $871~^{\circ}$C~\cite{Omega:2004}. The variation of the temperature in the environment surrounding the thermocouple is known to be much greater than that of the wire uncertainty. Expanded uncertainties as high as 20~\% for upper layer temperatures measured by a 1~mm bare-bead type K thermocouple have been reported by NIST researchers~\cite{Blevins:1999,Pitts:2003}. Small diameter thermocouples were used during these experiments to limit the impact of radiative heating and cooling. The estimated relative expanded uncertainty associated with the temperature measurements is $\pm$15~\%.

\subsubsection{Heat Flux Gauges}
Total heat flux measurements were made with water-cooled Schmidt-Boelter gauges. The manufacturer reports a $\pm$3~\% calibration expanded uncertainty for these devices~\cite{Medtherm:2003}. Results from an international study on total heat flux gauge calibration and response demonstrated that the relative expanded uncertainty of a Schmidt-Boelter gauge is typically $\pm$8~\%~\cite{Pitts:2006}.

\subsubsection{Gas Sampling}
The gas sampling system has a relative expanded uncertainty of $\pm$1~\% when compared with span gas volume fractions~\cite{Bundy:2007}. However, according to a study by Lock~et~al.~\cite{Lock:1}, the non-uniformity and movement of exhaust gases contribute to an estimated relative expanded uncertainty of $\pm$12~\%.

\subsubsection{Bi-Directional Probes}
Bi-directional probes were used to measure gas velocity in doorways. Bryant~\cite{Bryant:FSJ2009} estimates that the relative expanded uncertainty of this measurement ranges from $\pm$14~\% to $\pm$22~\%.

\subsubsection{Heat Release Rate}
A rotary type gas meter was used to measure the volume flow rate of propane into the gas burners. The manufacturer reports a relative standard uncertainty of $\pm$2~\%~\cite{Romet:2014}. A volumetric flow rate was calculated from the gas meter volume readings and used in conjunction with the heat of combustion of propane to calculate the heat release rate of the fire. The total expanded uncertainty for the heat release rate obtained from this method is estimated to be $\pm12$~\%.

@@

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\label{chap:acknowledgments}
The authors would like to thank Daniel Madrzykowski, Roy McLane, Kristopher Overholt, and Keith Stakes for their contributions in preparing and conducting these experiments.
